---
title: "Working with disk-backed mif objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with disk-backed mif objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

set.seed(126)
library(spatialTIME)
library(dplyr)

# Everything below writes into a temporary directory that is removed when the
# vignette finishes. In real use you would point these at a project directory.
store_dir <- file.path(tempdir(), "spatialTIME-vignette")
dir.create(store_dir, showWarnings = FALSE, recursive = TRUE)
```

# When the spatial data stops fitting

The `mif` object described in the [introduction](intro.html) holds every sample's
spatial data in memory, as a list of data frames. For TMA cores that is the right
thing to do. A core is a couple of thousand cells, the five cores shipped with this
package come to 12,782 cells in total, and the whole object is a few megabytes.

Whole slide images are a different problem. A single slide can be a million cells,
and a cohort of them does not fit. The numbers from the cohort this feature was
built for: 283 slides, 342,267,952 cells, 21 columns per cell. As R data frames that
is **34.2 GB**. On a 38.7 GB machine the `mif` is 89% of memory before any
computation starts, which means `ripleys_k(workers = 12)` cannot fork a single
worker — the run fails before it does any work.

The thing that makes this fixable is that the metrics do not need most of the data
most of the time. `ripleys_k()` reads a sample id, four bounding-box columns and
however many marker columns you asked for. On a 21-column HALO export that is 4
columns of 21, and on a real export with 51 columns it is worse. The other 17 are
carried in memory, in the parent process, for the entire run, and never read.

So the spatial slot can be a manifest of files instead of a list of data frames.
Each worker then reads only the columns it needs, for the one sample it is handling,
and the parent holds an index rather than a cohort. On those 283 slides the same
data is **2.19 GB** of parquet against 34.2 GB in memory.

## This is not a speed-up

Worth saying plainly, because it is easy to assume otherwise. Disk-backing does not
make the metrics faster, and on a single small sample it is slightly slower — there
is a file read that was not there before.

What it changes is *which* memory cost you pay. For Ripley's K the per-worker close
pair list from the engine overtakes the spatial frame at around `max(r_range) = 30`,
and at the default `r_range = seq(0, 100, 1)` it is roughly twelve times the frame.
The frame is not what makes a large run fail. The parent's copy of the whole cohort
is, and that is a fixed cost you cannot trade against `workers`. Disk-backing
removes it; `big` is the argument for bounding the pair list.

If your data fits in memory, keep it in memory.

# Writing and reading a store

`mif_to_disk()` takes an ordinary `mif` and a directory to write it to, and returns
a `mif` that reads from that directory.

```{r, write}
x <- create_mif(clinical_data = example_clinical %>%
                  mutate(deidentified_id = as.character(deidentified_id)),
                sample_data = example_summary %>%
                  mutate(deidentified_id = as.character(deidentified_id)),
                spatial_list = example_spatial,
                patient_id = "deidentified_id",
                sample_id = "deidentified_sample")

store <- file.path(store_dir, "cohort.mif")
xd <- mif_to_disk(x, store)

xd
```

`path` has no default. A store can be tens of gigabytes, so it is always written
where you say it should be and never somewhere chosen for you — the same reasoning
as `sigma` in `split_tissue()`.

The result is still a `mif`, with the same six slots, and the spatial slot still
answers to `length()`, `names()` and `[[`:

```{r, slots}
class(xd)
names(xd)
length(xd$spatial)
names(xd$spatial)
dim(xd$spatial[["TMA3_[9,K].tif"]])
```

That last line reads one file off disk. Iteration works too, so code written against
an in-memory `mif` does not need rewriting:

```{r, iterate}
sapply(xd$spatial, nrow)
```

There is a catch worth knowing before you use that on a real cohort. `lapply()` and
everything built on it materialise their input before applying anything, so
iterating the slot reads **every** sample — which is the cost disk-backing exists to
avoid. Iterating over `names()` instead keeps one sample resident at a time:

```{r, iterate-bounded}
vapply(names(xd$spatial), function(s) nrow(xd$spatial[[s]]), numeric(1))
```

Both give the same answer here, on five cores. On 283 slides the first reads 34 GB
and the second reads one slide at a time.

The slot itself is an index:

```{r, index}
xd$spatial
format(object.size(xd$spatial), units = "KB")
format(object.size(x$spatial), units = "MB")
```

The store is a plain directory you can inspect, copy, archive and move:

```{r, layout}
list.files(store, recursive = FALSE)
list.files(file.path(store, "spatial"))
```

`manifest.json` records the format version, the id columns, and each sample's file,
row count, byte size and column schema. Paths inside it are relative, so moving or
copying the whole directory does not break it.

`open_mif()` reads one back:

```{r, reopen}
y <- open_mif(store)
identical(names(y$spatial), names(xd$spatial))
```

Opening validates before returning. Every file is re-probed and compared against
the manifest — existence, byte size, row count and column names — so a truncated,
half-written or partially copied store fails immediately, with a message naming the
sample, rather than somewhere in the middle of a long computation. It costs about a
millisecond per sample, because a parquet footer read does not touch the data.

`collect_mif()` is the inverse of `mif_to_disk()`, named after `dplyr::collect()`
for the same reason: it is where lazily-referenced data gets materialised.

```{r, collect}
back <- collect_mif(xd)
class(back$spatial)
all.equal(back$spatial, x$spatial, check.attributes = FALSE)
```

On a cohort that does not fit in memory, collecting all of it defeats the point.
`samples` takes a subset, which is how you pull a few samples out to look at them:

```{r, collect-subset}
two <- collect_mif(xd, samples = c("TMA3_[9,K].tif", "TMA1_[3,B].tif"))
names(two$spatial)
sapply(two$spatial, nrow)
```

# Running the metrics

Nothing changes. Every function in the package takes a disk-backed `mif` the same
way it takes an in-memory one.

```{r, metrics}
mnames <- c("CD3..Opal.570..Positive", "CD8..Opal.520..Positive")

xd <- ripleys_k(mif = xd, mnames = mnames, r_range = seq(0, 100, 10),
                edge_correction = "translation", permute = FALSE,
                workers = 1, overwrite = TRUE)

head(xd$derived$univariate_Count[, c("deidentified_sample", "Marker", "r",
                                     "Observed K", "Exact CSR")])
```

The results are the same numbers. Not approximately — the same:

```{r, parity}
x <- ripleys_k(mif = x, mnames = mnames, r_range = seq(0, 100, 10),
               edge_correction = "translation", permute = FALSE,
               workers = 1, overwrite = TRUE)

all.equal(xd$derived$univariate_Count, x$derived$univariate_Count)
```

That is a tested contract rather than an aspiration. The in-memory path ignores the
column projection entirely, which means a projection that forgot a column a metric
actually uses would fail only on the disk path — so the package carries a parity
test comparing the two for every metric, at one and two workers, with and without
permutation.

Permuted results match too. Seeds are drawn once in the parent process and handed
to the workers, so a run depends only on your `set.seed()` and not on `workers`:

```{r, perm-parity}
set.seed(42)
a <- ripleys_k(xd, mnames = mnames[1], r_range = seq(0, 50, 10), permute = TRUE,
               num_permutations = 10, workers = 1, overwrite = TRUE)
set.seed(42)
b <- ripleys_k(x, mnames = mnames[1], r_range = seq(0, 50, 10), permute = TRUE,
               num_permutations = 10, workers = 1, overwrite = TRUE)

all.equal(a$derived$univariate_Count, b$derived$univariate_Count)
```

Derived results are written back to the store as they are computed, so they survive
a reopen:

```{r, derived-persist}
z <- open_mif(store)
names(z$derived)
identical(z$derived$univariate_Count, xd$derived$univariate_Count)
```

# Pointing at files you already have

If your data is already parquet, `create_mif()` accepts a named character vector of
file paths instead of a list of data frames. Nothing is copied.

```{r, reference}
files <- list.files(file.path(store, "spatial"), full.names = TRUE)
names(files) <- names(xd$spatial)

ref <- create_mif(clinical_data = example_clinical %>%
                    mutate(deidentified_id = as.character(deidentified_id)),
                  sample_data = example_summary %>%
                    mutate(deidentified_id = as.character(deidentified_id)),
                  spatial_list = files,
                  patient_id = "deidentified_id",
                  sample_id = "deidentified_sample")

ref$spatial
```

This is *reference mode*, and the print-out says so: there is no store root, because
the files are yours and live wherever you put them. Each file must hold exactly one
sample — every metric uses the convex hull of all of a file's cells as the
observation window, so a file with two samples in it would give both the wrong
window. If the vector is unnamed, the names are read from the `sample_id` column and
a file holding more than one distinct value is rejected.

The trade-off is that nothing may write to those files. A function that adds a
per-cell column has no business modifying your primary data, so `split_tissue()`
refuses on a reference-mode `mif` and tells you to make a store first.

# Subsetting

`subset_mif()` needs somewhere to put the result, for the same reason
`mif_to_disk()` does: the subset of a cohort too large for memory usually is too.

```{r, subset}
sub <- subset_mif(xd, classifier = "Classifier.Label", level = "Tumor",
                  markers = mnames, path = file.path(store_dir, "tumor.mif"))

cells <- function(m) vapply(names(m$spatial),
                            function(s) nrow(m$spatial[[s]]), numeric(1))

data.frame(all = cells(xd), tumor_only = cells(sub))
```

It writes sample by sample and releases each one, so peak memory stays at a single
sample regardless of cohort size.

# Checking that the files are still what you think

`open_mif()` validates at open time, but two situations are not covered by that. A
reference-mode `mif` has no manifest, so there was never anything to validate. And
a long session can hold a store open for hours while the files underneath it are
replaced or regenerated.

Both leave the recorded row count stale, and that matters more than it sounds: every
mask in this package is positional, so a stale row count means markers attached to
the wrong cells and plausible, wrong numbers rather than an error. `verify_mif()`
re-probes everything and reports the first file that no longer matches.

It returns `TRUE` invisibly and errors otherwise, so it is safe to put at the top of
a script — wrap it in `isTRUE()` if you want to see the answer.

```{r, verify}
isTRUE(verify_mif(xd))
isTRUE(verify_mif(ref))
isTRUE(verify_mif(x))     # in-memory: nothing to check, passes trivially
```

That last line is the point of it passing trivially in memory: you do not have to
branch on the representation to call it.

# Adding columns

`split_tissue()` writes three per-cell columns, and on a store those go to a
per-sample *overlay* file rather than into the base parquet.

```{r, split}
xd <- split_tissue(xd, classifier = "Classifier.Label",
                   class1 = "Tumor", class2 = "Stroma",
                   sigma = 40, interface_width = 100, workers = 1)

list.files(file.path(store, "overlay"))
colnames(mif_spatial <- xd$spatial[["TMA3_[9,K].tif"]])[1:4]
table(xd$spatial[["TMA3_[9,K].tif"]]$refined_density_compartment)
```

Appending columns by rewriting would touch all 2.19 GB of a cohort to add about 12
MB of new data, so the base files are never rewritten. Reads cbind the overlay back
on, and overlay columns shadow base columns of the same name, which is what lets a
re-run with `overwrite = TRUE` replace its own columns instead of duplicating them.

# What survives a round trip, and what does not

Column names, column order, types, and factor levels — including unused levels and
`NA`s — all round-trip exactly. That last one matters because `split_tissue()`
writes two factor columns per cell.

**Row names do not.** They come back as `1:nrow`. A columnar file has nowhere to put
them, storing them would mean an extra column per sample for something no function
here reads, and `dplyr` resets them anyway. Non-default row names only arise from
having subset a data frame, so this is visible only if you compare with
`identical()` after doing that. Use `all.equal(check.attributes = FALSE)`, as above,
or reset them yourself.

A few column types are refused outright rather than stored badly, and the check runs
before anything is written:

```{r, bad-column, error = TRUE}
bad <- example_spatial[["TMA3_[9,K].tif"]][1:5, ]
bad$notes <- as.raw(1:5)

mif_to_disk(toy <- create_mif(
  clinical_data = data.frame(deidentified_id = "p1"),
  sample_data = data.frame(deidentified_id = "p1",
                           deidentified_sample = "TMA3_[9,K].tif"),
  spatial_list = list(`TMA3_[9,K].tif` = bad),
  patient_id = "deidentified_id", sample_id = "deidentified_sample"),
  file.path(store_dir, "will-not-exist.mif"))
```

`raw` is the one that justifies the check: parquet accepts it and hands back an
`integer`, silently. List columns come back as something that is not what you put
in. If you have a column like that, drop it before writing or keep the `mif` in
memory.

The spatial slot of a store is read-only. The assignment forms that look like they
should work on a named list will not, because underneath the slot is a vector of
file paths and assigning to it would quietly destroy the index:

```{r, readonly, error = TRUE}
xd$spatial[[1]] <- data.frame(a = 1)
```

Use `collect_mif()` to bring the data into memory if you need to modify it directly.

# One thing to avoid

Do not resize arrow's thread pool from inside a worker. `arrow::set_cpu_count()` or
`set_io_thread_count()` called in an `mclapply()` child deadlocks — the run hangs at
0% CPU with no error and no output, which is about the worst failure mode available.
Reading parquet in a child is fine; it is only the resize that hangs.

If you need to limit arrow's threads, set the count in the parent before calling a
metric and the workers will inherit it safely.

```{r, threads, eval = FALSE}
arrow::set_cpu_count(2)        # in the parent, before the metric
xd <- ripleys_k(xd, mnames = mnames, r_range = 0:100, workers = 6)
```

```{r, cleanup, include = FALSE}
unlink(store_dir, recursive = TRUE)
```
