---
title: "Introduction to spatialTIME"
header-includes:
  - usepackage{mathbbm}
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spatialTIME}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(spatialTIME)
```

# Create Multiplex ImmunoFlourescent (mif) object 

spatialTIME functions use a custom mif object which can be created using `create_mif`. 

```{r, create}

x <- create_mif(clinical_data = example_clinical,
                sample_data = example_summary,
                spatial_list = example_spatial,
                patient_id = "deidentified_id", sample_id = "deidentified_sample",
                clean_columns = TRUE)
x

```

Four slots are available in an `mif` object. 
*spatial* is a list of data frames with each data frame containing the spatial 
information for a single sample. 
If this list is unnamed, then names with be assigned based off the `image_tag` value.
*clinical* is a data frame that contains relevant patient level data - at a 
minimum this data frame should contain patient level IDs (one row per patient). 
*sample* is a data frame containing any sample level values and contain sample 
names and patient IDs at a minimum (one row per sample). 
An empty list for *derived* is created as a place to store results from the 
`spatialTIME` package within the object (*this it the temporary usage - need to implement R6 classes to automatically update the argument supplied to the function*).

# Plotting Cores

An individual plot for each core (each sample) is created. 
Plots can be assigned to an R object, such as within the empty `derived` 
slot and printed to a PDF if a file name is provided. 

```{r, plot}
mnames <- c("foxp3_opal_620_positive", "cd3_opal_570_positive", "cd8_opal_520_positive",
            "pd1_opal_650_positive", "pdl1_opal_540_positive")

mlabels <- c("FOXP3", "CD3", "CD8", "PD1", "PDL1")

x[["derived"]][["spatial_plots"]] <- plot_immunoflo(x, plot_title = "deidentified_sample", 
                                                    mnames = mnames, mlabels = mlabels, 
                                                    cell_type = "classifier_label")

x[["derived"]][["spatial_plots"]][[4]]
```

# Estimating the degree of spatial clustering

Ripley's $K$ measures the average number of neighboring cells across each cell, that is the average number of cells within a specified radius of a cell. Ripley's $K$ is computed as follows:

$$K(r) = \frac{1}{n}\sum_{i=1}^{n}w_{ij}\mathbbm{1}(d(x_i,x_j)\le r)$$,

where $r$ is the specified radius, $d(x_i,x_j)$ is the distance between the $i^{th}$ and $j^{th}$ cell, and $w_{ij}$ is the weights that are assigned for border corrections. The expected value of $K(r)$ is $\pi r^2$, thus $K$ is expected to grow quadratically with $r$.

There are several edge correction, our studies has included a small number of cells and we recommend using the isotropic or translational edge correction, as opposed to the 'border' edge correction. The main goal of the edge correction is to account for the fact that there are unobserved points outside of the region, and the assumption is that the location of these cells has the same distribution as the study region.

There are several methods for normalization with help improve the interpretation of Ripley's $K$. We include Besag's L which is computed by $L(r) = \sqrt{\frac{K(r)}{\pi}}$, and the expect value of $r$. Hence, $L$ is expected to grow proportionally with $r$. The expected value $K$ and $L$ both hinge on the assumption that the underlying point process follows the so called complete spatial randomness assumption which asserts that the locations of cells can neither form a regular patterns nor clusters.

Damage can occur to TMAs, due to how they are collect. This damage can lead to rips and tears in the TMA which results in regions where it appears that cells cannot be locations, which is not actually the case. Thus the theoretical estimate for CSR may not be accurate, to address this the cell locations can be permuted and the permutation distribution of $K$ or $L$ is a TMA specific measure of CSR. The `ripleys_k` function reports a permuted and theoretical estimate of CSR, the observed value for $K$ (` kestimation = TRUE`) or $L$ (`kestimation = FALSE`), and the full permutation distribution of $K$ or $L$ if `keep_perm_dis = TRUE`.

```{r, ripleys, warning = FALSE}

x[["derived"]][["ripleys"]] <- ripleys_k(mif = x, id = "deidentified_sample", 
                                         mnames = mnames[1], num_permutations = 1,
                                         edge_correction = 'translation', 
                                         kestimation = TRUE, keep_perm_dis = FALSE)
```

# Estimating the colocalization of cell types


```{r, bi_ripleys, warning=FALSE}

mnames_pairs <- list(list("cd3_foxp3", "cd3_cd8"),
                     list("cd3_foxp3", "cd8_pd1"))


mlabels_pairs <- list(list("CD3+ FOXP3+", "CD3+ CD8+"),
                     list("CD3+ FOXP3+", "CD8+ PD1+"))

x[["derived"]][["bi-ripleys"]] <- bi_ripleys_k(mif = x, id = "deidentified_sample", 
                                               mnames = mnames_pairs, num_permutations = 1,
                                               r_range = seq(0, 100, 50), edge_correction = "translation", 
                                               kestimation = TRUE, keep_perm_dis = FALSE,
                                               mlabels = mlabels_pairs)

x[["derived"]][["bi-ripleys"]] %>% head()
```


