---
title: "Introduction to spatialIF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spatialIF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(spatialIHC)
```

# Create Multiplex ImmunoFlourescent (mif) object 

spatialIF functions use a custom mif object which can be created using `create_mif`. 

```{r, create}

load("../data/example_tma.rda")

# names(example_tma)
# grabbing image tag names to make sample and clinical file -----
spatial_names <- lapply(example_tma, function(x) {x$image_tag[[1]]})
spatial_names <- unlist(spatial_names)
# spatial_names <- gsub("\\.tiff", "", spatial_names)

set.seed(8675309)
example_sample <- data.frame(image_tag = spatial_names,
                             patient_id = sample(c("patient_x", "patient_y", "patient_z"),
                                                 10, replace = TRUE))

example_clinical <- data.frame(patient_id = c("patient_x", "patient_y", "patient_z"),
                               covar_one = c("low", "high", "low"),
                               covar_two = rnorm(3))

x <- create_mif(clinical_data = example_clinical,
                sample_data = example_sample,
                spatial_list = example_tma)
x

```

Four slots are available in an `mif` object. 
*spatial* is a list of data frames with each data frame containing the spatial 
information for a single sample. 
If this list is unnamed, then names with be assigned based off the `image_tag` value.
*clinical* is a data frame that contains relevant patient level data - at a 
minimum this data frame should contain patient level IDs (one row per patient). 
*sample* is a data frame containing any sample level values and contain sample 
names and pateint IDs at a minimum (one row per sample). 
An empty list for *derived* is created as a place to store results from the 
`spatialIF` package within the object (*this it the temporary usage - need to implement R6 classes to automatically update the argument supplied to the function*).

# Plotting Cores

An individual plot for each core (each sample) is created. 
Plots can be assigned to an R object, such as within the empty `derived` slot and printed to a PDF if a file name is provided. 
 
```{r, plot}
# mnames <- c("FOXP3..Opal.620..Positive", "CD3..Opal.570..Positive", "CD8..Opal.520..Positive",
#             "PD1..Opal.650..Positive", "PDL1..Opal.540..Positive")
# mnames <- janitor::make_clean_names(mnames)
mnames <- c("foxp3_opal_620_positive", "cd3_opal_570_positive", "cd8_opal_520_positive",
            "pd1_opal_650_positive", "pdl1_opal_540_positive")

mlabels <- c("FOXP3", "CD3", "CD8", "PD1", "PDL1")

x[["derived"]][["spatial_plots"]] <- plot_immunoflo(x, plot_title = "image_tag", mnames = mnames, mlabels = mlabels, cell_type = "classifier_label")

x[["derived"]][["spatial_plots"]][[2]]
```

# Univariate Ripleys K/L

CHRIS can you add some summary information here on calculating univariate ripley's k and what each of the output measures are and when you would suggest to use each. 

```{r, ripleys}

x[["derived"]][["ripleys"]] <- ripleys_k(x, id = "image_tag", mnames = mnames,
                                         csr_calculation = "observed")

x[["derived"]][["ripleys2"]] <- ripleys_k(x, id = "image_tag", mnames = mnames,
                                          csr_calculation = "permutation", 
                                          num_permutations = 5)

```

## Bivariate Ripleys K/L

```{r, bi_ripleys}

mnames_pairs <- list(list("cd3_foxp3", "cd3_cd8"),
                     list("cd3_foxp3", "cd8_pd1"))

x[["derived"]][["bi-ripleys"]] <- bi_ripleys_k(x, id = "image_tag", mnames = mnames_pairs,
                                               csr_calculation = "observed")

x[["derived"]][["bi-ripleys2"]] <- bi_ripleys_k(x, id = "image_tag", mnames = mnames_pairs,
                                                csr_calculation = "permutation",
                                                num_permutations = 5)

```
