---
title: "Using SpatialExperiment Objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spatialexperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spatialTIME)
```

Multiplex single-cell imaging datasets

We'll be using data from the [VectraPolarisData Bioconductor package](https://www.bioconductor.org/packages/release/data/experiment/vignettes/VectraPolarisData/inst/doc/VectraPolarisData.html) (Wrobel, J and Ghosh, T, 2022). This provides two datasets from the Vectra Polaris imaging system in the form of SpatialExperiment objects. [See this vignette for more details](http://juliawrobel.com/MI_tutorial/MI_VectraPolarisData.html).

```{r}
if (!("VectraPolarisData" %in% installed.packages())) {
  BiocManager::install("VectraPolarisData")
}
```

## About the data 

In this example we will be using data from a [human ovarian cancer study](https://pubmed.ncbi.nlm.nih.gov/34615692/) (Steinhart et al, 2021). 

```{r}
# Ovarian cancer example
ovarian <- VectraPolarisData::HumanOvarianCancerVP()
```

For more details on the `SpatialExperiment` class see this [Bioconductor Vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html) and [preprint](https://www.biorxiv.org/content/10.1101/2021.01.27.428431v3).

We can inspect what this type of object looks like typically:

```{r}
ovarian
```

The `colData` entry has a lot of our marker information as well as other annotation. Let's specifically look at the `sample_id` and some of the markers we will be using.

This is how we look at the colData in this set. 

```{r}
colData(ovarian)[1:5, 1:5]
```

We can look at the sample IDs doing this. 

```{r}
# this is how we can extract sample IDs
head(ovarian$sample_id)
```

There are also marker columns we will need for plotting They all start with `phenotype_`

```{r}
head(ovarian$phenotype_cd19)
```

Now to use these data with spatialTIME package for analyses we will need to convert it to the MIF (multiplex immunofluorescence) format. For SpatialExperiment objects specifically we can do this using the `spatial_exp_to_mif()` function. For all other types of data you can use the `create_mif()` function. 

There are three required arguments in this function to create our mif object. 
1) The `SpatialExperiment` object we are using. 
2) The name of the column is that links the colData with the spatialCoord data. In this case it is call "sample_id".
3) The names of the columns that contain any markers we'd like to use for the analysis that are included in the data as column names

```{r}
# These are the genes we will be using as markers that exist in these data. 
markers = c("phenotype_cd68", "phenotype_cd3", "phenotype_cd8")
```

Now we can pass all of this into the `spatial_exp_to_mif` function to create our `mif` object for use with `spatialTIME`.

```{r}
ova_mif <- spatial_exp_to_mif(
  spatial_exp = ovarian,
  patient_id = "sample_id",
  markers = markers,
  cols_to_keep = "tissue_category"
)
```


These objects come with so many columns of data that spatialTIME doesn't need so many will be dropped from the object. But if there is a particular column you will need for labeling purposes, you can name it in the `cols_to_keep` variable. 

## Plot it 

Now we can use our `mif` object to make a set of immunofluoresence plots, one for each sample in the set.  

```{r}
ova_mif <- plot_immunoflo(ova_mif,
  plot_title = "sample_id",
  mnames = markers,
  xloc = "x",
  yloc = "y"
)
```

We can view individual plots by pulling them out using the indexing notation. 
Where `[[1]]` would be the first sample in the list and `[[3]]`, the third sample in the list, etc. 

````{r}
ova_mif$derived$spatial_plots[[3]] +
  ggplot2::coord_equal()
```

We could build on to this plot using `ggplot2`  and tidyverse code. 

## Subset mIF to Single Compartment 

We can subset our mif object to a certain compartment. Note this is why when we made our object we `cols_to_keep = "tissue_category"` because we are using that labeling now. 

```{r}
# filter the mif object to be only the tumor compartment
ova_mif_tumor <- subset_mif(ova_mif,
  classifier = "tissue_category",
  level = "Tumor",
  markers = markers
)
```

Now that we have ran the subset to tumor only, let's check a spatial data frame for levels in `tissue_category`

```{r}
# check a spatial data frame for levels in tissue_category
ova_mif_tumor$spatial[[1]]$tissue_category %>% unique()
```

Success! We only have cells in a tumor compartment.
Summary of the samples has also been updated to reflect new compartment.

```{r}
names(ova_mif_tumor$sample)
```


## Ripley's K 


Now we can use Ripley's K stat to quantify our spatial clustering. 

```{r}
# takes time to process all samples
ova_mif_tumor <- ripleys_k(ova_mif_tumor,
  mnames = markers,
  r_range = 0:100,
  edge_correction = "translation", # translational edge correction
  permute = FALSE, # allows for exact estimation of complete spatial ransomness, or CSR
  workers = 1, # depending on number of spatial samples and ram, may cause crashes if overallocated
  big = 50000, # really only needed (keep larger than largest spatial data) when computer has ram limitations otherwise would avoid, causes non-perfect estimation of CSR due to rounding
  xloc = "x",
  yloc = "y",
  overwrite = TRUE
)
```

Now that we've run Ripley's K, let's plot `phenotype_cd3` to view t-cell clustering

```{r}
ova_mif_tumor$derived$univariate_Count %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Degree of Clustering Exact`, group = sample_id),
    alpha = 0.5
  )
```

We can see that most samples have a positive 'degree of clustering' meaning that the T cells in the most samples are clustering closer together than expected if they were randomly distributed around the samples. 
We also see that there are 202 rows that have missing values, i.e. the spatial files didn't have enough positive cells for a marker.

## Nearest Neighbor G

Alternatively we can use Nearest Neighbor G to examine clustering which identifies the probability of encountering a neighbor cell. 

```{r}
ova_mif_tumor <- NN_G(
  mif = ova_mif_tumor,
  mnames = markers,
  r_range = 0:100,
  num_permutations = 10, # NN G(r) requires permutations to estimate CSR
  edge_correction = "rs", # reduced sampling edge correction
  keep_perm_dis = FALSE, # whether to keep results from each permutation
  workers = 1,
  overwrite = TRUE,
  xloc = "x",
  yloc = "y"
)

```

Now let's plot `phenotype_cd3` again but with NNG results to view t-cell clustering.

```{r}
ova_mif_tumor$derived$univariate_NN %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Degree of Clustering Permutation`, group = sample_id),
    alpha = 0.5
  )
```

We can see NN G(r) is much more 'curvy' than Ripley's K.
values above 0 mean clustering more than expected if the T cells were random
we also can vaguely see that there is a bit of a hill between 0 and 25, indicating tight clusters of CD3, 
then outside of that they either fall to 0 or maintain their degree of clustering
NN G(r) (observed) tells the proportion of T cells with a neighbor at or below a measured radius.

```{r}
ova_mif_tumor$derived$univariate_NN %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Observed G`, group = sample_id),
    alpha = 0.5
  )
```

By plotting the raw NN G(r) value, we see that it grows and approaches 1, at which point all T cells in the sample has another T cell within that radius. Or 75% of T cells have another T cell within whatever radius.
The reason we take the difference between observed and permuted (Degree of Clustering Permutation) is to account for sample-specific technical artifacts like holes or even necrotic reagions that violate what Ripley's K and NN G assume of the sample this difference is then what we compare between samples

## Picking a Radius - Ripley's K 

Unless diving into complicated statistics with functional data analysis (analysis of the entire shape of the RIpley's K or NN G curve), researchers have to pick a radius that they would like to associate with some feature of the samples like patient survival, stage, treatment response, or even disease subtypes

Best way found to do this is plot all curves for a marker and find a radius that's small enough to biologically meaningful while large enough to capture variance in the data

Let's plot phenotype_cd3 to view t-cell clustering again to take a look.

```{r}
ova_mif_tumor$derived$univariate_Count %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Degree of Clustering Exact`, group = sample_id),
    alpha = 0.5
  )
```

For this data, we might pick a radius of 50 pixels because after that most values don't drastically change, so we can subset and move forward.

```{r}
analysis_dat <- ova_mif_tumor$derived$univariate_Count %>%
  filter(Marker == "phenotype_cd3", r == 50) %>% # filter ripley's k output to CD3 cells and radius of 50
  full_join(
    ova_mif_tumor$clinical, # merge with clinical (on left side)
    .
  ) # ripleys k on right
head(analysis_dat)
```

## Performing Association with Survival

For simplicity we can dichotimize the clustering values

```{r}
analysis_dat <- analysis_dat %>%
  mutate(
    cluster_group = ifelse(`Degree of Clustering Exact` >= median(`Degree of Clustering Exact`, na.rm = TRUE), # remembering that 2 samples didn't have enough T cells so remove NA
      "High Clustering",
      "Low Clustering"
    ), # if above the median, high, else low
    cluster_group = factor(cluster_group, levels = c("Low Clustering", "High Clustering"))
  ) %>% # convert to factor to control reference
  filter(!is.na(cluster_group)) # must remove extra rows that do not have clustering group
# create formula with columns in analysis_data
form1 <- formula(Surv(survival_time, as.numeric(death)) ~ age_at_diagnosis + stage + grade + cluster_group)
# fit the model with the data
model1 <- coxph(form1, data = analysis_dat)
# view the fit
summary(model1)
```

We see that stage 4 has moderately worse survival than stage 1 (reference) but not significant based on exp(coef)
hazard ratio (exp[-coef]) = 7.75, 95% confidence interval (0.869, 69.116)
we also see that high clustering has a moderately worse survival than the low clustering
hazard ratio (exp[-coef]) = 1.55, 95% confidence interval (0.966, 2.482)

We can also see whether or not adding the clustering to the model formula significantly improves the model fit to the data
This must be done with a nested formula, i.e. removing or adding a column to the formula to assess impact of the addition or removal
Here we just remove the `clustering_group` and fit another model
Create formula with columns in analysis_data

```{r}
form2 <- formula(Surv(survival_time, as.numeric(death)) ~ age_at_diagnosis + stage + grade)
```

Now we can fit the model with the data. 

```{r}
# fit the model with the data
model2 <- coxph(form2, data = analysis_dat)
```

To identify the significance, we can perform a likelihood ratio test
```{r}
lrt <- lrtest(
  model2,
  model1
)

# Let's print it out
lrt
```

It doesn't appear that adding in the spatial information signficantly improves the survival model fit for the data

## Session info 

```{r}
sessionInfo()
```
