---
title: "Using SpatialExperiment Objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spatialexperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spatialTIME)
```

Multiplex single-cell imaging datasets

We'll be using data from the [VectraPolarisData Bioconductor package](https://www.bioconductor.org/packages/release/data/experiment/vignettes/VectraPolarisData/inst/doc/VectraPolarisData.html) (Wrobel, J and Ghosh, T, 2022). This provides two datasets from the Vectra Polaris imaging system in the form of SpatialExperiment objects. [See this vignette for more details](http://juliawrobel.com/MI_tutorial/MI_VectraPolarisData.html).

```{r}
if (!("VectraPolarisData" %in% installed.packages())) {
  BiocManager::install("VectraPolarisData")
}
```

## About the data 

In this example we will be using data from a [human ovarian cancer study](https://pubmed.ncbi.nlm.nih.gov/34615692/) (Steinhart et al, 2021). 

```{r}
# Ovarian cancer example
ovarian <- VectraPolarisData::HumanOvarianCancerVP()
```

For more details on the `SpatialExperiment` class see this [Bioconductor Vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html) and [preprint](https://www.biorxiv.org/content/10.1101/2021.01.27.428431v3).

We can inspect what this type of object looks like typically:

```{r}
ovarian
```

The `colData` entry has a lot of our marker information as well as other annotation. Let's specifically look at the `sample_id` and some of the markers we will be using.

This is how we look at the colData in this set. 

```{r}
colData(ovarian)
```

We can look at the sample IDs doing this. 

```{r}
# this is how we can extract sample IDs
head(ovarian$sample_id)
```

There are also marker columns we will need for plotting They all start with `phenotype_`

```{r}
head(ovarian$phenotype_cd19)
```

Now to use these data with spatialTIME package for analyses we will need to convert it to the MIF (multiplex immunofluorescence) format. For SpatialExperiment objects specifically we can do this using the `spatial_exp_to_mif()` function. For all other types of data you can use the `create_mif()` function. 

There are three required arguments in this function to create our mif object. 
1) The `SpatialExperiment` object we are using. 
2) The name of the column is that links the colData with the spatialCoord data. In this case it is call "sample_id".
3) The names of the columns that contain any markers we'd like to use for the analysis.

```{r}
ova_mif <- spatial_exp_to_mif(
  spatial_exp = ovarian,
  patient_id = "sample_id",
  markers = c("phenotype_cd68", "phenotype_cd3", "phenotype_cd8")
)
```

## Plot it 

```{r}
ova_mif <- plot_immunoflo(ova_mif,
  plot_title = "sample_id",
  mnames = markers,
  xloc = "x",
  yloc = "y"
)
# view a sample
ova_mif$derived$spatial_plots[[3]] +
  coord_equal()
```

```{r}
# do similar with ggplot
ova_mif$spatial[[3]] %>% # select the same sample
  mutate(Other = ifelse(rowSums(across(!!markers)) == 0, # if none of the markers are positive make "Other" 1 else 0
    1,
    0
  )) %>%
  select(x, y, tissue_category, !!markers, Other) %>% # select only needed columns
  pivot_longer(c(!!markers, Other), # take cell-type columns and convert to one column
    names_to = "Marker", # called "Marker"
    values_to = "Positive"
  ) %>% # and column contents a new columns called "Positive
  filter(Positive == 1) %>% # select just values of 1 (less duplicates)
  ggplot() + # prep plotting object
  geom_point(
    data = . %>%
      filter(Marker == "Other"), # only plot the background 'other' cells
    aes(
      x = x, y = y, # with locations
      shape = tissue_category
    )
  ) + # and shape of tissue
  geom_point(
    data = . %>%
      filter(Marker != "Other"), # plot now positive cells
    aes(
      x = x, y = y, # with locations
      shape = tissue_category, # tissue
      color = Marker
    )
  ) + # and color by cell type (without other)
  coord_equal() + # make x and y same scale
  theme_classic() + # clean it up
  scale_y_reverse() + # make same orientation as image (flipped)
  ggplot2::scale_shape_manual(NULL,
    values = c(3, 16),
    drop = FALSE
  ) # make the shapes the right shape
```

## Subset mIF to Single Compartment 

```{r}
# filter the mif object to be only the tumor compartment
ova_mif_tumor <- subset_mif(ova_mif,
  classifier = "tissue_category",
  level = "Tumor",
  markers = markers
)
# check a spatial data frame for levels in tissue_category
ova_mif_tumor$spatial[[1]]$tissue_category %>% unique()
```

great, only cells in a tumor compartment
summary of the samples has also been updated to reflect new compartment

## Ripley's K 

```{r}
# takes time to process all samples
ova_mif_tumor <- ripleys_k(ova_mif_tumor,
  mnames = markers,
  r_range = 0:100,
  edge_correction = "translation", # translational edge correction
  permute = FALSE, # allows for exact estimation of complete spatial ransomness, or CSR
  workers = 1, # depending on number of spatial samples and ram, may cause crashes if overallocated
  big = 50000, # really only needed (keep larger than largest spatial data) when computer has ram limitations otherwise would avoid, causes non-perfect estimation of CSR due to rounding
  xloc = "x",
  yloc = "y",
  overwrite = TRUE
)
```

plot phenotype_cd3 to view t cell clustering

```{r}
ova_mif_tumor$derived$univariate_Count %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Degree of Clustering Exact`, group = sample_id),
    alpha = 0.5
  )
```

can see that most samples have a positive 'degree of clustering'
meaning that the T cells in the most samples are clustering closer together than expected
if they were randomly distributed around the samples. 
we also see that there are 202 rows that have missing values, i.e. the spatial files didn't have enough positive cells for a marker

## Nearest Neighbor G

```{r}
ova_mif_tumor <- NN_G(
  mif = ova_mif_tumor,
  mnames = markers,
  r_range = 0:100,
  num_permutations = 10, # NN G(r) requires permutations to estimate CSR
  edge_correction = "rs", # reduced sampling edge correction
  keep_perm_dis = FALSE, # whether to keep results from each permutation
  workers = 1,
  overwrite = TRUE,
  xloc = "x",
  yloc = "y"
)
# plot phenotype_cd3 to view t cell clusering
ova_mif_tumor$derived$univariate_NN %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Degree of Clustering Permutation`, group = sample_id),
    alpha = 0.5
  )
```

We can see NN G(r) is much more 'curvy' than Ripley's K
values above 0 mean clustering more than expected if the T cells were random
we also can vaguely see that there is a bit of a hill between 0 and 25, indicating tight clusters of CD3, 
then outside of that they either fall to 0 or maintain their degree of clustering
NN G(r) (observed) tells the proportion of T cells with a neighbor at or below a measured radius

```{r}
ova_mif_tumor$derived$univariate_NN %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Observed G`, group = sample_id),
    alpha = 0.5
  )
```

By plotting the raw NN G(r) value, we see that it grows and approaches 1, at which point all T cells in the sample has another T cell within that radius. Or 75% of T cells have another T cell within whatever radius.
The reason we take the difference between observed and permuted (Degree of Clustering Permutation) is to account for sample-specific technical artifacts like holes or even necrotic reagions that violate what Ripley's K and NN G assume of the sample this difference is then what we compare between samples

## Picking a Radius - Ripley's K 

unless diving into complicated statistics with functional data analysis (analysis of the entire shape of the RIpley's K or NN G curve)
researchers have to pick a radius that they would like to associate with some feature of the samples like patient survival,
stage, treatment response, or even disease subtypes

best way found to do this is plot all curves for a marker and find a radius thats small enough to biologically meaningful while
large enough to capture variance in the data
plot phenotype_cd3 to view t cell clusering

```{r}
ova_mif_tumor$derived$univariate_Count %>%
  filter(Marker == "phenotype_cd3") %>%
  ggplot() +
  geom_line(aes(x = r, y = `Degree of Clustering Exact`, group = sample_id),
    alpha = 0.5
  )
```


For this data, we might pick a radius of 50 pixels because after that most values don't drastically change, so we can subset and move forward.

```{r}
analysis_dat <- ova_mif_tumor$derived$univariate_Count %>%
  filter(Marker == "phenotype_cd3", r == 50) %>% # filter ripley's k output to CD3 cells and radius of 50
  full_join(
    ova_mif_tumor$clinical, # merge with clinical (on left side)
    .
  ) # ripleys k on right
head(analysis_dat)
```

## Performing Association with Survival

For simplicity we can dichotomize the clustering values

```{r}
analysis_dat <- analysis_dat %>%
  mutate(
    cluster_group = ifelse(`Degree of Clustering Exact` >= median(`Degree of Clustering Exact`, na.rm = TRUE), # remembering that 2 samples didn't have enough T cells so remove NA
      "High Clustering",
      "Low Clustering"
    ), # if above the median, high, else low
    cluster_group = factor(cluster_group, levels = c("Low Clustering", "High Clustering"))
  ) %>% # convert to factor to control reference
  filter(!is.na(cluster_group)) # must remove extra rows that do not have clustering group
# create formula with columns in analysis_data
form1 <- formula(Surv(survival_time, as.numeric(death)) ~ age_at_diagnosis + stage + grade + cluster_group)
# fit the model with the data
model1 <- coxph(form1, data = analysis_dat)
# view the fit
summary(model1)
```

We see that stage 4 has moderately worse survival than stage 1 (reference) but not significant based on exp(coef)
hazard ratio (exp[-coef]) = 7.75, 95% confidence interval (0.869, 69.116)
we also see that high clustering has a moderately worse survival than the low clustering
hazard ratio (exp[-coef]) = 1.55, 95% confidence interval (0.966, 2.482)

We can also see whether or not adding the clustering to the model formula significantly improves the model fit to the data
This must be done with a nested formula, i.e. removing or adding a column to the formula to assess impact of the addition or removal
Here we just remove the clustering_group and fit another model
Create formula with columns in analysis_data

```{r}
form2 <- formula(Surv(survival_time, as.numeric(death)) ~ age_at_diagnosis + stage + grade)
# fit the model with the data
model2 <- coxph(form2, data = analysis_dat)

# to identify the significance, we can perform a likelihood ratio test
lrt <- lrtest(
  model2,
  model1
)
lrt
# doesn't appear that adding in the spatial information signficantly improves the survival model fit for the data
```


## Session info 
```{r}
sessionInfo()
```
